Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-06-14 SIMPaper/SIMPaper.Rmd
added data availability statement

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "Neural network-based survival methods have been developed to model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling framework with flexible architectures. Our method naturally accounts for censoring, does not require additional hyperparameters and uses binary cross-entropy as the loss function. Using a novel sampling scheme and data augmentation, we incorporate time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a given moment and estimates the hazard function. We compare the performance of CBNN to survival methods based on regression and neural networks in two simulations and two real data applications. We report two time-dependent metrics for each model. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time and outperforms the competing neural network approaches. For a simple simulation with an exponential hazard model, CBNN outperforms the other neural network methods. For a complex simulation, which highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, CBNN outperforms all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. We highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of survival outcomes. An R package is available at [https://github.com/Jesse-Islam/cbnn](https://github.com/Jesse-Islam/cbnn)."

@@

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855 to RS. We would like to thank Dr. James Meigs, Project Leader of these awards, for his support and helpful discussions. We would also like to Thank Dr. James Hanley for his support and discussions while extending the case-base methodology.

>>>>

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
abstract: "Neural network-based survival methods can model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling framework with flexible architectures. Our method naturally accounts for censoring and does not require method specific hyperparameters. Using a novel sampling scheme and data augmentation, we incorporate time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a given moment and estimates the hazard function. We compare the performance of CBNN to survival methods based on regression and neural networks in two simulations and two real data applications. We report two time-dependent metrics for each model. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time and outperforms the competing neural network approaches. For a simple simulation with an exponential hazard model, CBNN outperforms the other neural network methods. For a complex simulation, which highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, CBNN outperforms all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. We highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of survival outcomes. An R package is available at [https://github.com/Jesse-Islam/cbnn](https://github.com/Jesse-Islam/cbnn)."

@@

# Data and code availability statement {-}

The pre-processed data for the SUPPORT case study can be found at [https://github.com/jaredleekatzman/DeepSurv/tree/master/experiments/data/support](https://github.com/jaredleekatzman/DeepSurv/tree/master/experiments/data/support). The pre-processed data for the METABRIC case study can be found at [https://github.com/jaredleekatzman/DeepSurv/tree/master/experiments/data/metabric](https://github.com/jaredleekatzman/DeepSurv/tree/master/experiments/data/metabric). The data is accessed using [https://github.com/havakv/pycox](https://github.com/havakv/pycox). The code for this manuscript and its analyses can be found at [https://github.com/Jesse-Islam/cbnnManuscript](https://github.com/Jesse-Islam/cbnnManuscript). The software package making CBNN easier to use can be found at [https://github.com/Jesse-Islam/cbnn](https://github.com/Jesse-Islam/cbnn).

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855 to RS. We would like to thank Dr. James Meigs, Project Leader of these awards, for his support and helpful discussions. The work was also supported as part of the Congressionally Directed Medical Research Programs (CDMRP) award W81XWH-17-1-0347. We would also like to Thank Dr. James Hanley for his support and discussions while extending the case-base methodology.

### 2022-06-14 SIMPaper/SIMPaper.tex
added data availability statement

\revised{2022-06-01}

\accepted{2022-07-01}

@@

\abstract{Neural network-based survival methods have been developed to
model data-driven covariate interactions. While these methods have led
to better predictive performance than regression-based approaches, they
cannot model both time-varying interactions and complex baseline
hazards. To address this, we propose Case-Base Neural Networks (CBNN) as
a new approach that combines the case-base sampling framework with
flexible architectures. Our method naturally accounts for censoring,
does not require additional hyperparameters and uses binary
cross-entropy as the loss function. Using a novel sampling scheme and
data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. We highlight the benefit of combining
case-base sampling with deep learning to provide a simple and flexible
modeling framework for data-driven, time-varying interaction modeling of
survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855
to RS. We would like to thank Dr.~James Meigs, Project Leader of these
awards, for his support and helpful discussions. We would also like to
Thank Dr.~James Hanley for his support and discussions while extending
the case-base methodology.

>>>>

@@

\abstract{Neural network-based survival methods can model data-driven
covariate interactions. While these methods have led to better
predictive performance than regression-based approaches, they cannot
model both time-varying interactions and complex baseline hazards. To
address this, we propose Case-Base Neural Networks (CBNN) as a new
approach that combines the case-base sampling framework with flexible
architectures. Our method naturally accounts for censoring and does not
require method specific hyperparameters. Using a novel sampling scheme
and data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. We highlight the benefit of combining
case-base sampling with deep learning to provide a simple and flexible
modeling framework for data-driven, time-varying interaction modeling of
survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

\hypertarget{data-and-code-availability-statement}{%
\section*{Data and code availability
statement}\label{data-and-code-availability-statement}}
\addcontentsline{toc}{section}{Data and code availability statement}

The pre-processed data for the SUPPORT case study can be found at
\url{https://github.com/jaredleekatzman/DeepSurv/tree/master/experiments/data/support}.
The pre-processed data for the METABRIC case study can be found at
\url{https://github.com/jaredleekatzman/DeepSurv/tree/master/experiments/data/metabric}.
The data is accessed using \url{https://github.com/havakv/pycox}. The
code for this manuscript and its analyses can be found at
\url{https://github.com/Jesse-Islam/cbnnManuscript}. The software
package making CBNN easier to use can be found at
\url{https://github.com/Jesse-Islam/cbnn}.

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855
to RS. We would like to thank Dr.~James Meigs, Project Leader of these
awards, for his support and helpful discussions. The work was also
supported as part of the Congressionally Directed Medical Research
Programs (CDMRP) award W81XWH-17-1-0347. We would also like to Thank
Dr.~James Hanley for his support and discussions while extending the
case-base methodology.
